### 2022-04-01 abstractESSLLI21/abstractESSLLI1.Rmd
Adds bitly

\begin{center}
\Large \textbf{A bayesian method of cosine-based word2vec bias estimation}
\end{center}

@@

A considerable amount of literature exists on  bias detection and mitigation in NLP models, especially word2vec embeddings, which represent words as  vectors of real numbers (see e.g. @bolukbasi2016man and @Gonen2019lipstick and references therein). The most common method used compares  cosine similarity between words from protected groups and attributes that are considered to be stereotypical or harmful in some way, and this method will be in our focus.

@@

\footnotesize \begin{align}
\mathsf{s}(t,A,B)  = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,
\mathsf{bias}(A,B)  = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
} \tag{WEAT}
\end{align}
\normalsize
\noindent @Caliskan2017semanticsBiases show that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings. @Lauscher2019multidimensional extend the methodology to a multilingual and cross-lingual setting; a similar methodology is employed by @Garg2018years, who use word embeddings trained on corpora from different decades. @Manzini2019blackToCriminal modify WEAT to a multi-class setting, introducing Mean Average Cosine (MAC) similarity as a measure of bias (in fact, in the paper  they report distances rather than similarities). Let $T = \{t_1, \dots, t_k\}$ be a class of protected word embeddings, and let each $A_j\in A$ be a set of attributes stereotypically associated with a protected word). Then:

@@

\noindent
\begin{tabular}{ll}
\begin{minipage}[c]{0.35\linewidth}
\scalebox{0.65}{
\begin{tabular}{cccccc}\toprule
$X$ to $A$  & $X$ to $B$ & $Y$ to $A$ & $Y$ to $B$ & $\sigma$  & WEAT \\
\midrule
0 & 0 & 0 & 0 & 0.05 & 1.82 \\
0 & 0 & 0 & 0 & 0.001 & -1.93\\
.1 & 0 & 0 & .1 & 0.05 & 1.49 \\
.1 & 0 & 0 & .1 & .1 & 1.22\\
\bottomrule
\end{tabular}
}
\end{minipage}& \begin{minipage}[c]{0.6\linewidth}

We have randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$. One simulation draws normally distributed values  for two situations in which the underlying mean similarities are in fact equal to zero, in two other cases  the means are different, with a range of choices of common standard deviation  accross the cases (code available upon request).

\end{minipage}
\end{tabular}

The following observations are worthwhile. (1) For points randomly drawn from distributions where there is no difference in means the calculated effect size can easily be  1.82, whereas the largest effect size reported by @Caliskan2017semanticsBiases is 1.81. (2) For samples from distributions where the means are different, the (absolute) effect sizes can easily be lower than in the first two simulations.  As Figure 2. illustrates,  quite some uncertainty is involved, far more than what systematically low mean-based p-values reported in the papers might suggest. Part of the problem is random variation unaccounted for in the original approach (see Figures 3 and 4 for an example), and part of the problem is that    non-negligible changes in effect size can result from a shift in the standard deviation of the original process, because with the decrease of standard deviation the numerator in (WEAT) decreases.

To improve on the situation, we  build  Bayesian models to estimate the biases involved using the raw datapoints, actually using control groups,  distinguishing the connection types,  and taking the uncertainty involved seriously.
For the general impact of being associated, we build  models using hamiltonian Monte Carlo (STAN) according to the following specification (linear categorical model and the priors):

@@

\begin{center}
\begin{figure}[!htb]\centering
   \begin{minipage}{0.48\textwidth}
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
ggplot(religion, aes(x =  cosineDistance, fill = connection, color = connection))+geom_density(alpha=0.6,size = .2)+theme_tufte()+labs(title  = "Figure 1.", subtitle = "Empirical distribution of cosine distances (religion, Reddit)")+ scale_fill_manual(values = c("orangered4","chartreuse4", "skyblue", "gray"))+scale_x_continuous(breaks = seq(0.3,1.5, by = 0.1))+xlab("cosine distance")+ scale_color_manual(values = c("orangered4","chartreuse4","skyblue","gray"))
```
   \end{minipage}
   \begin {minipage}{0.48\textwidth}
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
s <- function (table){ mean(table$A) - mean(table$B)}
biasesD2 <- numeric(10000)
for(i in 1:10000){
t1d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
t2d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
t3d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
t4d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))

factorD2 <- sd(c(s(t1d2),s(t2d2),s(t3d2),s(t4d2)))
numeratorD2 <-  mean(s(t1d2),s(t2d2)) - mean(s(t3d2),s(t4d2))
biasesD2[i] <- numeratorD2/factorD2
}

ggplot()+geom_histogram(aes(x=biasesD2, y = ..density..), alpha = 0.6, bins=50)+
  theme_tufte()+labs(title="Figure 2.", subtitle = "10k biases for different means  and sd =.05 (case 3 in the table)")+ xlab("bias")
```
   \end{minipage}

  \begin{minipage}{0.48\textwidth}
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%",message =FALSE, warning = FALSE}
library(tidyverse)
muslimWords <- c("imam","islam","mosque","muslim","quran")
muslim <- religion %>% filter(protectedWord %in% muslimWords)
muslimClass <- muslim %>% filter(protectedWord == "muslim")
neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
muslimVis <- rbind(muslimVis,neutralSample,humanSample)
#we plug in our visualisation script
source("../functions/visualisationTools.R")
#two arguments: dataset and protected word
visualiseProtected(muslimVis,"muslim")+labs(title = "Figure 3.",
subtitle = "Cosine distances for active attributes (musim, Reddit)")
```
\end{minipage}
   \begin {minipage}{0.48\textwidth}
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
library(tidyverse)
priestClass <- religion %>% filter(protectedWord == "priest")
neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
priestVis <- rbind(priestVis,neutralSample,humanSample)

#we plug in our visualisation script
source("../functions/visualisationTools.R")
#two arguments: dataset and protected word
visualiseProtected(priestVis,"priest")+labs(title = "Figure 4.",
subtitle = "Cosine distances for active attributes (priest, Reddit)")
```
   \end{minipage}

  \begin{minipage}{0.38\textwidth}
\includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
\end{minipage}
   \begin {minipage}{0.58\textwidth}
\includegraphics{../images/visReligionReddit.png}
   \end{minipage}
\end{figure}

\end{center}

\scriptsize

\vspace{-4mm}

>>>>

\begin{center}
\large \textbf{Taking uncertainty seriously}

\textbf{A  Bayesian approach to  word embedding bias estimation }
\end{center}

@@

A considerable amount of literature exists on  bias detection and mitigation in NLP models, especially word embeddings, which represent words as  vectors of real numbers (see e.g. @Bolukbasi2016man and @Gonen2019lipstick and references therein). The most common methods used compare cosine similarity or distance between words from protected groups and attributes that are considered to be stereotypical or harmful in some way. Many of them rely on pre-averaging, which we find problematic --- two such methods will be in our focus. We will argue that the existing measures of bias by pre-averaging oversimplify the picture and lead to false confidence. Our simulation results show that the mainstream interpretations are too hasty. Our Bayesian analysis with proper control groups suggests that (1) the word lists in use are too short, (2) associated and different groups behave similarly, which undermines the semantic interpretation of cosine similarity, and (3) our analysis at different levels of generality shows that there is more variety than one may initially think.

@@

\footnotesize \begin{align}
\mathsf{s}(t,A,B)  = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,
\mathsf{bias}(A,B)  = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
} \tag{WEAT}
\end{align}
\normalsize
\noindent @Caliskan2017semanticsBiases show that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings. @Lauscher2019multidimensional extend the methodology to a multilingual and cross-lingual setting; a similar methodology is employed by @Garg2018years, who use word embeddings trained on corpora from different decades. @Manzini2019blackToCriminal modify WEAT to a multi-class setting, introducing Mean Average Cosine (MAC) similarity distance as a measure of bias. Let $T = \{t_1, \dots, t_k\}$ be a class of protected word embeddings, and let each $A_j\in A$ be a set of attributes stereotypically associated with a protected word). Then:

@@

\begin{center}
\begin{tabular}{cccccc}\toprule
$X$ to $A$  & $X$ to $B$ & $Y$ to $A$ & $Y$ to $B$ & $\sigma$  & WEAT \\
\midrule
0 & 0 & 0 & 0 & 0.05 & 1.82 \\
0 & 0 & 0 & 0 & 0.001 & -1.93\\
.1 & 0 & 0 & .1 & 0.05 & 1.49 \\
.1 & 0 & 0 & .1 & .1 & 1.22\\
\bottomrule
\end{tabular}
\end{center}

To illustrate, we have randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$. One simulation draws normally distributed values  for two situations in which the underlying mean similarities are in fact equal to zero, in two other cases  the means are different, with a range of choices of common standard deviation  across the cases (see the table). 
The following observations are worthwhile. For points randomly drawn from distributions where there is no difference in means the calculated effect size can easily be  1.82, whereas the largest effect size reported by by Caliskan is 1.81. For samples from distributions where the means are different, the (absolute) effect sizes can easily be lower than in the first two simulations. Moreover, 10k simulations with more realistic word list sizes (normal distributions with equal means, standard deviation=0.08, 16 protected words and 16 predicates) show that high scores are easy to get even if there is no difference in the generating model (Figure 2, code available upon request). This indicates that the low p-value reported in the papers are highly misleading.

To improve on the situation, we  build  Bayesian models to estimate the biases involved using the raw data points, actually using control groups,  distinguishing the connection types,  and taking the uncertainty involved seriously.
For the general impact of being associated, we build  models using Hamiltonian Monte Carlo (STAN) according to the following specification (linear categorical model and the priors):

@@

\includegraphics[width=8cm]{violinReligion.pdf}
\includegraphics[width=8cm]{simulations.pdf}

\includegraphics[width=8cm]{muslim.pdf}
\includegraphics[width=8cm]{priest.pdf}

\includegraphics[width=8cm]{../images/religionCoeffs.jpeg}
\includegraphics[width=8cm]{../images/visReligionReddit.png}

<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"} -->
<!-- ggplot(religion, aes(x =  cosineDistance, fill = connection, color = connection))+geom_density(alpha=0.6,size = .2)+theme_tufte()+labs(title  = "Figure 1.", subtitle = "Empirical distribution of cosine distances (religion, Reddit)")+ scale_fill_manual(values = c("orangered4","chartreuse4", "skyblue", "gray"))+scale_x_continuous(breaks = seq(0.3,1.5, by = 0.1))+xlab("cosine distance")+ scale_color_manual(values = c("orangered4","chartreuse4","skyblue","gray")) -->
<!-- ``` -->

<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"} -->
<!-- s <- function (table){ mean(table$A) - mean(table$B)} -->
<!-- biasesD2 <- numeric(10000) -->
<!-- for(i in 1:10000){ -->
<!-- t1d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05)) -->
<!-- t2d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05)) -->
<!-- t3d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05)) -->
<!-- t4d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05)) -->

<!-- factorD2 <- sd(c(s(t1d2),s(t2d2),s(t3d2),s(t4d2))) -->
<!-- numeratorD2 <-  mean(s(t1d2),s(t2d2)) - mean(s(t3d2),s(t4d2)) -->
<!-- biasesD2[i] <- numeratorD2/factorD2 -->
<!-- } -->

<!-- ggplot()+geom_histogram(aes(x=biasesD2, y = ..density..), alpha = 0.6, bins=50)+ -->
<!--   theme_tufte()+labs(title="Figure 2.", subtitle = "10k biases for different means  and sd =.05 (case 3 in the table)")+ xlab("bias") -->
<!-- ``` -->

<!-- # ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%",message =FALSE, warning = FALSE} -->
<!-- # library(tidyverse) -->
<!-- # muslimWords <- c("imam","islam","mosque","muslim","quran") -->
<!-- # muslim <- religion %>% filter(protectedWord %in% muslimWords) -->
<!-- # muslimClass <- muslim %>% filter(protectedWord == "muslim") -->
<!-- # neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5) -->
<!-- # humanSample <- sample_n(filter(muslimClass,connection == "human"), 5) -->
<!-- # muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human") -->
<!-- # muslimVis <- rbind(muslimVis,neutralSample,humanSample) -->
<!-- # #we plug in our visualisation script -->
<!-- # source("../functions/visualisationTools.R") -->
<!-- # #two arguments: dataset and protected word -->
<!-- # visualiseProtected(muslimVis,"muslim")+labs(title = "Figure 3.", -->
<!-- # subtitle = "Cosine distances for active attributes (musim, Reddit)") -->
<!-- # ``` -->

 <!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"} -->
<!-- library(tidyverse) -->
<!-- priestClass <- religion %>% filter(protectedWord == "priest") -->
<!-- neutralSample <- sample_n(filter(priestClass,connection == "none"), 5) -->
<!-- humanSample <- sample_n(filter(priestClass,connection == "human"), 5) -->
<!-- priestVis <- priestClass %>% filter(connection != "none" & connection !="human") -->
<!-- priestVis <- rbind(priestVis,neutralSample,humanSample) -->

<!-- #we plug in our visualisation script -->
<!-- source("../functions/visualisationTools.R") -->
<!-- #two arguments: dataset and protected word -->
<!-- visualiseProtected(priestVis,"priest")+labs(title = "Figure 4.", -->
<!-- subtitle = "Cosine distances for active attributes (priest, Reddit)") -->
<!-- ``` -->

\pagebreak
## References {-}
\scriptsize

### 2022-04-01 abstractESSLLI21/abstractESSLLI1.tex
Adds bitly

\usepackage{multirow}
\usepackage{multicol}

\usepackage{graphicx}
\usepackage{longtable}
\usepackage{ragged2e}
\usepackage{etex}
%\usepackage{yfonts}
\usepackage{marvosym}
\usepackage[notextcomp]{kpfonts}
\usepackage{nicefrac}
\newcommand*{\QED}{\hfill \footnotesize {\sc Q.e.d.}}
\usepackage{floatrow}

@@

\usepackage{geometry}
 \geometry{a4paper,left=20mm,top=15mm,bottom = 20mm, right = 20mm}

@@

A considerable amount of literature exists on bias detection and
mitigation in NLP models, especially word2vec embeddings, which
represent words as vectors of real numbers (see e.g. {[}1{]} and {[}4{]}
and references therein). The most common method used compares cosine
similarity between words from protected groups and attributes that are
considered to be stereotypical or harmful in some way, and this method
will be in our focus.

@@

\footnotesize \begin{align}
\mathsf{s}(t,A,B)  = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,
\mathsf{bias}(A,B)  = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
} \tag{WEAT}
\end{align}

\normalsize
\noindent {[}2{]} show that significant biases---thus measured---
similar to the ones discovered by IAT can be discovered in word
embeddings. {[}5{]} extend the methodology to a multilingual and
cross-lingual setting; a similar methodology is employed by {[}3{]}, who
use word embeddings trained on corpora from different decades. {[}6{]}
modify WEAT to a multi-class setting, introducing Mean Average Cosine
(MAC) similarity as a measure of bias (in fact, in the paper they report
distances rather than similarities). Let \(T = \{t_1, \dots, t_k\}\) be
a class of protected word embeddings, and let each \(A_j\in A\) be a set
of attributes stereotypically associated with a protected word). Then:

@@

\noindent

\begin{tabular}{ll}
\begin{minipage}[c]{0.35\linewidth}
\scalebox{0.65}{
\begin{tabular}{cccccc}\toprule
$X$ to $A$  & $X$ to $B$ & $Y$ to $A$ & $Y$ to $B$ & $\sigma$  & WEAT \\
\midrule
0 & 0 & 0 & 0 & 0.05 & 1.82 \\
0 & 0 & 0 & 0 & 0.001 & -1.93\\
.1 & 0 & 0 & .1 & 0.05 & 1.49 \\
.1 & 0 & 0 & .1 & .1 & 1.22\\
\bottomrule
\end{tabular}
}
\end{minipage}& \begin{minipage}[c]{0.6\linewidth}

We have randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$. One simulation draws normally distributed values  for two situations in which the underlying mean similarities are in fact equal to zero, in two other cases  the means are different, with a range of choices of common standard deviation  accross the cases (code available upon request).

\end{minipage}
\end{tabular}

The following observations are worthwhile. (1) For points randomly drawn
from distributions where there is no difference in means the calculated
effect size can easily be 1.82, whereas the largest effect size reported
by {[}2{]} is 1.81. (2) For samples from distributions where the means
are different, the (absolute) effect sizes can easily be lower than in
the first two simulations. As Figure 2. illustrates, quite some
uncertainty is involved, far more than what systematically low
mean-based p-values reported in the papers might suggest. Part of the
problem is random variation unaccounted for in the original approach
(see Figures 3 and 4 for an example), and part of the problem is that
non-negligible changes in effect size can result from a shift in the
standard deviation of the original process, because with the decrease of
standard deviation the numerator in (WEAT) decreases.

To improve on the situation, we build Bayesian models to estimate the
biases involved using the raw datapoints, actually using control groups,
distinguishing the connection types, and taking the uncertainty involved
seriously. For the general impact of being associated, we build models
using hamiltonian Monte Carlo (STAN) according to the following
specification (linear categorical model and the priors):

@@

\begin{center}
\begin{figure}[!htb]\centering
   \begin{minipage}{0.48\textwidth}

\begin{center}\includegraphics[width=1\linewidth]{abstractESSLLI1_files/figure-latex/unnamed-chunk-1-1} \end{center}
   \end{minipage}
   \begin {minipage}{0.48\textwidth}

\begin{center}\includegraphics[width=1\linewidth]{abstractESSLLI1_files/figure-latex/unnamed-chunk-2-1} \end{center}
   \end{minipage}

  \begin{minipage}{0.48\textwidth}

\begin{center}\includegraphics[width=1\linewidth]{abstractESSLLI1_files/figure-latex/unnamed-chunk-3-1} \end{center}
\end{minipage}
   \begin {minipage}{0.48\textwidth}

\begin{center}\includegraphics[width=1\linewidth]{abstractESSLLI1_files/figure-latex/unnamed-chunk-4-1} \end{center}
   \end{minipage}

  \begin{minipage}{0.38\textwidth}
\includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
\end{minipage}
   \begin {minipage}{0.58\textwidth}
\includegraphics{../images/visReligionReddit.png}
   \end{minipage}
\end{figure}

\end{center}

@@

\vspace{-4mm}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\hypertarget{ref-bolukbasi2016man}{}%
\CSLLeftMargin{{[}1{]} }
\CSLRightInline{Tolga Bolukbasi, Kai-Wei Chang, James Zou, Venkatesh
Saligrama, and Adam Kalai. 2016. Man is to computer programmer as woman
is to homemaker? Debiasing word embeddings. Retrieved from
\url{http://arxiv.org/abs/1607.06520}}

>>>>

\usepackage{multirow}
\usepackage{multicol}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{ragged2e}
\usepackage{etex}
%\usepackage{yfonts}
\usepackage{marvosym}
\usepackage[notextcomp]{kpfonts}
\usepackage{nicefrac}
\newcommand*{\QED}{\hfill \footnotesize {\sc Q.e.d.}}
\usepackage{floatrow}

@@

\usepackage{geometry}
 \geometry{a4paper,left=35mm,top=20mm,bottom = 20mm, right = 35mm}

@@

A considerable amount of literature exists on bias detection and
mitigation in NLP models, especially word embeddings, which represent
words as vectors of real numbers (see e.g. {[}1{]} and {[}4{]} and
references therein). The most common methods used compare cosine
similarity or distance between words from protected groups and
attributes that are considered to be stereotypical or harmful in some
way. Many of them rely on pre-averaging, which we find problematic ---
two such methods will be in our focus. We will argue that the existing
measures of bias by pre-averaging oversimplify the picture and lead to
false confidence. Our simulation results show that the mainstream
interpretations are too hasty. Our Bayesian analysis with proper control
groups suggests that (1) the word lists in use are too short, (2)
associated and different groups behave similarly, which undermines the
semantic interpretation of cosine similarity, and (3) our analysis at
different levels of generality shows that there is more variety than one
may initially think.

@@

\footnotesize \begin{align}
\mathsf{s}(t,A,B)  = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,
\mathsf{bias}(A,B)  = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
} \tag{WEAT}
\end{align} \normalsize \noindent {[}2{]} show that significant
biases---thus measured--- similar to the ones discovered by IAT can be
discovered in word embeddings. {[}5{]} extend the methodology to a
multilingual and cross-lingual setting; a similar methodology is
employed by {[}3{]}, who use word embeddings trained on corpora from
different decades. {[}6{]} modify WEAT to a multi-class setting,
introducing Mean Average Cosine (MAC) similarity distance as a measure
of bias. Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word
embeddings, and let each \(A_j\in A\) be a set of attributes
stereotypically associated with a protected word). Then:

@@

\begin{center}
\begin{tabular}{cccccc}\toprule
$X$ to $A$  & $X$ to $B$ & $Y$ to $A$ & $Y$ to $B$ & $\sigma$  & WEAT \\
\midrule
0 & 0 & 0 & 0 & 0.05 & 1.82 \\
0 & 0 & 0 & 0 & 0.001 & -1.93\\
.1 & 0 & 0 & .1 & 0.05 & 1.49 \\
.1 & 0 & 0 & .1 & .1 & 1.22\\
\bottomrule
\end{tabular}
\end{center}

To illustrate, we have randomly drawn protected classes,
\(X=\{t_1,t_2\}\) and \(Y=\{t_3,t_4\}\), and two five-element attribute
sets \(A\) and \(B\). One simulation draws normally distributed values
for two situations in which the underlying mean similarities are in fact
equal to zero, in two other cases the means are different, with a range
of choices of common standard deviation across the cases (see the
table). The following observations are worthwhile. For points randomly
drawn from distributions where there is no difference in means the
calculated effect size can easily be 1.82, whereas the largest effect
size reported by by Caliskan is 1.81. For samples from distributions
where the means are different, the (absolute) effect sizes can easily be
lower than in the first two simulations. Moreover, 10k simulations with
more realistic word list sizes (normal distributions with equal means,
standard deviation=0.08, 16 protected words and 16 predicates) show that
high scores are easy to get even if there is no difference in the
generating model (Figure 2, code available upon request). This indicates
that the low p-value reported in the papers are highly misleading.

To improve on the situation, we build Bayesian models to estimate the
biases involved using the raw data points, actually using control
groups, distinguishing the connection types, and taking the uncertainty
involved seriously. For the general impact of being associated, we build
models using Hamiltonian Monte Carlo (STAN) according to the following
specification (linear categorical model and the priors):

@@

\includegraphics[width=8cm]{violinReligion.pdf}
\includegraphics[width=8cm]{simulations.pdf}

\includegraphics[width=8cm]{muslim.pdf}
\includegraphics[width=8cm]{priest.pdf}

\includegraphics[width=8cm]{../images/religionCoeffs.jpeg}
\includegraphics[width=8cm]{../images/visReligionReddit.png}

\hypertarget{references}{%
\subsection*{References}\label{references}}
\addcontentsline{toc}{subsection}{References}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\hypertarget{ref-Bolukbasi2016man}{}%
\CSLLeftMargin{{[}1{]} }
\CSLRightInline{Tolga Bolukbasi, Kai-Wei Chang, James Y. Zou, Venkatesh
Saligrama, and Adam Kalai. 2016. Man is to computer programmer as woman
is to homemaker? Debiasing word embeddings. \emph{CoRR} abs/1607.06520,
(2016). Retrieved from \url{http://arxiv.org/abs/1607.06520}}
